---
title: "Tidyverse manipulation"
author: "Lamia Benyamine"
date: "06/04/2024"
format: html
editor: visual
---

# *Task 1:* Conceptual Questions

1.  If your working directory is myfolder/homework/, what relative path would you specify to get the file located at myfolder/MyData.csv?

> "../MyData.csv"

2.  What are the major benefits of using R projects?

> Keeping all the files related to a project in one spot helps when other people need to replicate the results.

3.  What is git and what is github?

> Git is a version control software and Github is the online hosting service for Git projects

4.  What are the two main differences between a tibble and a data.frame?

> A tibble is a form of a data frame, but it has features that make it easier to use. For instance, when printing, a data frame shows the entire data while a tibble only shows the first 10 columns and column descriptions. Also, when subsetting the data, a tibble will return a tibble, but a data frame can sometimes return a vector or list.

5.  Rewrite the following nested function call using baseR’s chaining operator: arrange(filter(select(as_tibble(iris), starts_with("Petal"), Species), Petal.Length \< 1.55), Species)

> iris \|\>\
> as_tibble() \|\>\
> select(starts_with("Petal"), Species) \|\>\
> filter(Petal.Length \< 1.55) \|\>\
> arrange(Species)

6.  What is meant by long format data and wide format data? Which do we generally prefer for statistical analysis?

> Wide format is a pivot view where each row can have more than one observation because each column is a different variable. Whereas in the long format, each row consists of one observation and one variable in each column and this is the preferred view for data manipulation.

# *Task 2:* Reading Delimited Data

## Glass Data

Read in the Glass data from URL

```{r}
library(readr)
glass_tb <- read_csv("https://www4.stat.ncsu.edu/~online/datasets/glass.data",
                     col_names = c("Id", "RI", "Na", "Mg", "Al", "Si", "K", "Ca", "Ba", "Fe", "Type_of_glass"))
glass_tb
```

Add the glass type description and filter on Fe \< 0.2 and Glass type

```{r glass_type}
library(dplyr)
glass_tb |>
  mutate(Type_of_glass = ifelse(glass_tb$Type_of_glass == 1, "building_windows_float_processed", 
                                ifelse(glass_tb$Type_of_glass == 2, "building_windows_non_float_processed", 
                                       ifelse(glass_tb$Type_of_glass == 3,"vehicle_windows_float_processed", 
                                              ifelse(glass_tb$Type_of_glass == 4, "vehicle_windows_non_float_processed", 
                                                     ifelse(glass_tb$Type_of_glass == 5, "containers", 
                                                            ifelse(glass_tb$Type_of_glass == 6,"tableware", 
                                                                   ifelse(glass_tb$Type_of_glass == 7, "headlamps", "")))))))) |>
  collect() |>
  filter(Fe < 0.2, Type_of_glass %in% c("tableware", "headlamps"))
```

## Yeast Data

Read in the Yeast data from URL

```{r}
yeast_tb <- read_delim("https://www4.stat.ncsu.edu/~online/datasets/yeast.data", 
           delim = "  ",
           col_names = c("seq_name", "mcg", "gvh", "alm", "mit", "erl", "pox", "vac", "nuc", "class"))
yeast_tb
```

Manipulate the data by removing some columns and adding summary columns

```{r}
yeast_tb |>
  select(-1, -9) |>
  group_by(class) |>
   mutate(across(where(is.numeric), list(mean = mean, median=median), .names = "{.col}_{.fn}"))
```

# *Task 3:* Combining Excel and Delimited Data

## White Wine

Import white wine data

```{r}
white_tb <- readxl::read_excel("white-wine.xlsx", sheet = 1)
white_tb
```

Read in second sheet of wine data to get column names

```{r}
wine_names <- readxl::read_excel("white-wine.xlsx", sheet = 2)
colnames(white_tb) <- wine_names[[1]]

#add a new variable for type of wine
white_tb$type = "white"
```

## Red Wine

Import red wine data and follow the same maniupulations done for the white wine

```{r}
red_tb <- readr::read_delim("red-wine.csv",
                            delim = ";" ,
                            skip = 1,
                            col_names = wine_names[[1]])
red_tb$type = "red"
red_tb
```

## Combine Data

```{r}
wine_tb <- dplyr::bind_rows(white_tb, red_tb)
```

Filter, sort, and add summarized columns to the data

```{r}
wine_tb |>
  filter(quality > 6.5, alcohol < 132) |>
  arrange(desc(quality)) |>
  select(contains("acid"), "alcohol", "type", "quality") |>
  group_by(quality) |>
   mutate(mean_alcohol = mean(alcohol), sd_alcohol = sd(alcohol))
```

# *Task 4:* Database Practice

Connect to database and view tables

```{r}
library(DBI)
lah <-dbConnect(RSQLite::SQLite(), "lahman.db")
dbListTables(lah)
```

Return the 2015 data from Teams using dplyr and SQL

```{r}
#dplyr
tbl(lah, "Teams") |>
  filter(yearID == 2015) |>
  show_query()

#SQL
tbl(lah, sql(
  "SELECT `Teams`.*
    FROM `Teams`
    WHERE (`yearID` = 2015.0)")
)
```

Return all players in the hall of fame

```{r}
tbl(lah, "HallofFame") |>
  filter(inducted == "Y") |>
  select("playerID", "yearID", "category")
```

Combine the Hall of fame table with People to get full names

```{r}
hof <- left_join(tbl(lah, "HallofFame") |> filter(inducted == "Y"), 
          tbl(lah, "People"),
          by = join_by(playerID)) |>
  collect() |>
  select(playerID, yearID, category, nameFirst, nameLast)
hof
```

Create summary statistics for all managers

```{r}
mgrs <- tbl(lah, "Managers") |>
  select(playerID, G, W, L) |>
  group_by(playerID) |>
  summarize(G_managed = sum(G, na.rm = TRUE),
  Total_W = sum(W, na.rm = TRUE),
  Total_L = sum(L, na.rm = TRUE)) |>
  collect() |>
  mutate(w_L_percent = round((Total_W / G_managed) * 100,2)) |>
  arrange(desc(w_L_percent))
mgrs
```

Return managers of teams that are in the hall of fame

```{r}
inner_join(hof, mgrs)
```

Disconnect from database

```{r}
dbDisconnect(lah)
```
